When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-04-06 doc/paper/vpd_et_paper.tex
pierre's comments done

Biases between the theory and observations are similar for ENF, WSA,
CRO and SAV. At low VPD the theory and the observations match
well. However, at high VPD (upper 10-20\% of observations) the theory
is biased slightly towards positive response as compared to the
observations (Figure \ref{test_sign}). For ENF this slight bias could
be explained by a negative bias in $g_1$. However, for WSA, CRO, and
SAV the observations in the highest VPD bins exhibit a downturn
towards more negative response, which cannot be captured by the
functional form of the sign term (Figure
\ref{idealized_sign}). Therefore, to explain the observations one of
the variables in the sign term must change at extreme VPD (upper
10-20\%). The most likely candidate is $uWUE$, which we have assumed
constant to meet our goal of developing intuition for leading order
behavior, but might be expected to decrease at extremely low SWC,
which should be correlated with high VPD. The theory for EBF exhibits
similar limitations as for ENF, WSA, and SAV, except a larger
portion of observations are biased negative as compared to the theory
($\approx$ 35\%, Figure \ref{test_sign}). However, in general, and
specifically for non-extreme VPD (VPD < $\approx$70-90th percentiles), the theory matches the observations
for the tree and savannah plant types.

@@

While the above discussion shows that our theory has some limitations
when applied to some PFTs, especially for grasslands, it does well for
DBF, MF and CSH PFTs, and captures the response at non-extreme VPD
(VPD < $\approx$ 70th-90th percentiles) for
ENF, EBF, CRO, WSA, and SAV. In general, the leading order behavior
observed in the data is captured by the theory. Departures between our
theory and observations, specifically at high VPD, could be explained
and conceptualized with shifts in $g_1$ and/or $uWUE$ due to
site-specific plant type variability (e.g. more arid-adapted ecosystems
at more arid sites), or temporal variability for some environmental
conditions (e.g. decreases in $uWUE$ at extremely low soil
moisture). To focus on general behavior and develop intuition for
PFT-scale response, we have ignored these sources of variability in
the present analysis. However, any site-specific or temporal plant
functional variability that can be conceptualized with shifts to $g_1$
and $uWUE$ can be analyzed with our framework. This opens a door to a
myriad future analysis in which plant behavior in anomalous conditions
can be explained and analyzed using Equation \ref{et}.

@@

Therefore, with the caveat that some of the signal presented in Figure
\ref{data_scatter} may in fact be noise, we can interpret the observed
distribution of ET response to VPD. In general, the observed response
matches the intuitive theory. ET response to VPD shifts towards
positive values as VPD increases (atmospheric demand dominating). CRO
exhibit the highest occurrence of positive ET response, and the
observations confirm that CRO are the most water intensive. CSH are
the most water conservative, with a strong negative response. DBF,
EBF, ENF, MF, SAV and WSA are also water conservative, but show some
occurrence of positive ET response to VPD, particularly at higher
observed VPD, as the theory predicts. GRA, while generally water
conservative, does not match the theory, with increasing frequency of
\textit{negative} response at high VPD. This is as expected, given
discussion in Section \ref{testing}.

@@

Indeed, for all PFTs, there is some slight dependence of
$\sigma \cdot uWUE$ on SWC, especially at low SWC (Figure
\ref{swc_boxplot}). The portion of observations for which our theory
is biased by SWC-limitations varies by PFT, due to the nonlinear
threshold at which soil moisture availability limits plant
function. For CRO, MF and DBF, soil water content only matters for
about the lowest 5\% of observations (each box is 5\% of observations
in Figure \ref{swc_boxplot}). And, CRO, relative to DBF, has a very
weak dependence on soil moisture, which is reflective of the high
likelihood that CRO sites are optimally irrigated and not water
stressed, suggesting that observed departures from the theory for CRO
are due to other factors (i.e. photosynthesis pathway) than soil
moisture. For ENF, EBF, SAV, and WSA, systematic SWC-induced biases in
$\sigma \cdot uWUE$ emerge in about the lowest 20-30\% of SWC
conditions, although large variability in the SWC-$\sigma \cdot uWUE$
relationship hampers interpretation for WSA. CSH presents a special
case: the limited number of sites (2) preclude comment on the
exact PFT-wide relationship.

@@

The observed dependence of $\sigma \cdot uWUE$ on SWC for GRA would
explain the deficiencies of our theory compared to the observations in
Figure \ref{idealized_sign}, specifically the trend back towards
negative soil moisture at high VPD. Aforementioned feedbacks between
the land surface and the atmosphere, which are not accounted for due
to our focus on the one-way response of the land surface to
atmospheric conditions, would cause high VPD to be correlated with low
SWC \citep[][]{Gentine_2016, Berg_2016}. So, at high VPD observations
of low SWC are more likely, and this low SWC causes a lower uWUE
(Figure \ref{swc_boxplot}). The lower uWUE at low SWC/high VPD then
leads to the observed downturn towards decreasing ET with increasing
VPD, and the deviation between our theory (based on a constant uWUE
assumption) and observations. It is also perhaps not coincidental that
the portion of $\sigma \cdot uWUE$ affected by low soil moisture
observations is similar to the portion of observations that to not
match our theory at high VPD. Indeed, this may suggest that for all PFTs
except for GRA, coupling and feedbacks between SWC, VPD and plant
function are relatively rare.

@@

The theory described in Section \ref{sign_term} indicates that for a
given $uWUE$ and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minima in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms  (e.g. aerodynamic
conductance, $\Delta$, $g_1$ and $uWUE$) are held fixed. This result warrants further
investigation, because to our knowledge no one has derived the
theoretical ecosystem-scale relationship between ET and VPD while
controlling for environmental conditions. In particular, from
personal communication, there is an apparent lack of consensus over
whether the shape of the ET-VPD curve should be concave upward (our
result) or concave downward. Given that understanding the ET-VPD
relationship of the one-way plant response is fundamental to
hypothesizing about any feedbacks between the land surface and the
atmosphere, we analyze why our derived ET-VPD relationship is concave
upward, particularly with respect to the exponent of VPD dependence in
$uWUE$ and the Medlyn unified stomatal conductance model.

There is a theoretical basis for the square root VPD dependence in
both the stomatal conductance model and $uWUE$ based on the assumption
that stomata behave to maximize carbon gain while minimizing water
loss, which observations also generally support \citep{Lloyd_1991,
  MEDLYN_2011, Lin_2015, Zhou_2014, Zhou_2015, Medlyn_2017}. However,
some purely empirical results which fit the exponent of the VPD
dependence to data have shown that it may vary slightly from 1/2,
suggesting that stomata, as well as ecosystem-scale quantities based
on stomata theory, may not always function optimally \citep{Zhou_2015,
  Lin_2018}. Specifically with regards to $uWUE$, one would not expect
that this ecosystem scale WUE quantity will respond to VPD exactly
analogously to stomata. Direct soil evaporation's contributions to ET
should shift the exponent of the VPD dependence, especially for
conditions of low GPP when we would expect a systematically larger
portion of direct soil evaporation contributions to ET, because we
would also expect lower amounts of transpiration. \citet{Zhou_2015}'s
results corroborate this: they found a mean empirically fit
exponential VPD dependence of 0.55, varying slightly from the
theoretically optimal value of of 0.5 for AmeriFlux
sites. Interpreting \citet{Lin_2018}'s results, which also show
variance in the empirical exponent of the VPD dependence of the
stomatal conductance model, is more difficult as \citet{Lin_2018} do
not handle GPP/A dependence of stomatal conductance in a directly
analogous manner to the optimal theory in \citet{MEDLYN_2011} and
\citet{Medlyn_2017}. Regardless, given that these recent results on
the relationship between VPD, GPP, and ET \citep{MEDLYN_2011,
  Zhou_2014, Zhou_2015, Medlyn_2017} form the backbone of our analysis
and are what allowed us to derive an explicit ET expression for the
first time (Equation \ref{et}), we will analyze if and how assumptions
about the exponent of the VPD dependence impacts the shape of the
ET-VPD dependence. By introducing $n$ and $m$ we can free our stomatal
conductance model from assumptions about VPD dependence:
\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_*}{VPD^m}\right) \frac{*WUE \; ET}{c_a \; VPD^n},
    \label{m_n}
  \end{equation}
\end{linenomath*}
where:
\[*WUE = \frac{GPP}{ET}VPD^n,\] and $g_*$ is a generic slope parameter
of units $VPD^m$. To determine how the exponent $n$ and $m$ alter the
shape of the ET-VPD dependence we find the roots of the second
derivative of ET, using Equation \ref{m_n} for stomatal conductance
($g_s$, with respect to VPD: \input{d2_solutions.tex} With this
result we've defined the family of curves separating concave up from
concave down ET solutions (Figure \ref{concave}). These curves are
only functions of the exponent of the VPD dependence and a quantity we
call non-dimensional VPD ($VPD^m/g*$). Several important
relations reveal themselves from Equation \ref{curves}:

@@

\begin{enumerate}
  \item For optimal behavior (n, m = 1/2) the ET-VPD curve will be
    concave up regardless of the magnitude of the plant constants
    $g_1$ and $uWUE$. Therefore, the general concave up nature of our
    results, given an assumption of optimal behavior, is insensitive
    to plant type.
  \item For all physically possible exponents of VPD dependence ($n,
    m$), whether the solution is concave up or concave down does not
    depend on $uWUE$.
  \item In general, increasing the exponent of VPD dependence
    increases the likelihood of a concave up result. Additionally, as
    the exponent of VPD dependence increases from the optimum value of
    1/2, whether the curve is concave upward or concave downward
    becomes a function of the plant specific slope parameter $g_*$,
    through non-dimensional VPD ($VPD^m/g_*$). Because the exponent of
    the VPD dependencies is capable of altering the fundamental shape
    of ET-VPD dependence, future research investment in understanding
    the exact VPD dependence of stomatal conductance, and further
    reconciliation of empirical and theoretical stomatal and
    ecosystem behavior should be prioritized.
\end{enumerate}
While it is possible that in the future some other form of
VPD dependence is derived, at present \cite{MEDLYN_2011} and
\cite{Zhou_2014} firmly established n=m=1/2 as the most likely
candidate given current theory and empirical data. Additionally, we
argue that a concave up result matches physical intuition more than a
concave down result. Plants must maintain nutrient and sugar transport
through the phloem and xylem. To accomplish this, stomata must remain
slightly open \citep{De_2013, Nikinmaa_2013, Ryan_2014}. Furthermore,
even if complete stomatal closure were possible, cuticular water loss
and [at the ecosystem-scale] direct soil evaporation are still sources
of ET which increase with VPD, independent of stomatal
closure. Therefore, in the limit as VPD becomes large and we assume
plants are exercising all strategies to reduce ET, any further increase
in VPD should result in an increase in ET through cuticular water loss
and/or direct soil evaporation. This inevitable transition from
conditions when stomata respond strongly to VPD to conditions when
stomata response is asymptoting towards full closure would cause a
concave up ET-VPD curve, which is matched by the theory. In short,
plant response becomes more limited as VPD increases, while
atmospheric demand monotonically increases with VPD, leading to the
intuitive result that atmospheric demand dominates plant response when
atmospheric demand is high.

@@

Lastly, our paper builds important intuition for how plants will
respond to VPD perturbations. We show that given optimal stomatal
function and fixed environmental conditions, the ET-VPD dependence is
theoretically concave up, with ET increasing with increasing VPD as
VPD increases past some critical value (Table \ref{vpd_crit}). However
future research should focus on fully understanding the form of
stomatal VPD dependence, as this result is sensitive to the exponent of
VPD dependence, which we currently believe is 1/2 \citep{MEDLYN_2011,
  Zhou_2014}. Our results are applicable to understanding the impact
of  expected increases in VPD
induced by global change.  Plant physiological responses to direct
CO$_2$ effects \citep[e.g.,][]{Swann_2016} receives more attention
than physiological response to indirect effects like increased
VPD. Here, we provide broad PFT-focused results showing a likely
decrease in ET in response to positive VPD perturbations (atmospheric
drying), which is consistent with recent observational analysis
\citep[e.g.,][]{Rigden_2017}. Feedbacks between the land and the
atmosphere may alter the net response to a long-timescale global VPD perturbation,
but our focus on the one way plant response to a VPD perturbation in
the atmospheric boundary layer is an important first step to
disentangling such feedbacks, both in observations and model
simulations of the present and future. By removing Penman Monteith's
dependence on implicit relationships between GPP, VPD, and ET, we
allow for explicit, accurate future analysis of plant-VPD feedbacks
in the atmospheric boundary layer (Equation \ref{et}). Our approach
can be extended to examine varying plant response to more nuanced
consideration of plant type and climate. Any plant physiological
heterogeneity or feedback that can be conceptualized with shifts in $g_1$
\citep[e.g.][]{Lin_2015, Medlyn_2017} and/or uWUE
\citep[e.g.][]{Zhou_2014} are representable within our framework, which
opens the door for a hierarchy of more sophisticated climate- and
plant-specific analysis of ET sensitivity to environmental variables
(including VPD). We argue that such simplified conceptual frameworks
are critical tools for disentangling land-atmosphere feedbacks at
various scales, from diurnal to seasonal and beyond, and to
characterize ET response in a warmer, atmospherically drier, and
enriched CO$_2$ world.

>>>>

Biases between the theory and observations are similar for ENF, WSA,
CRO and SAV. At low VPD the theory and the observations match
well. However, at high VPD (upper 10-20\% of observations) the theory
is biased slightly towards positive response as compared to the
observations (Figure \ref{test_sign}). For ENF this slight bias could
be explained by a negative bias in $g_1$. However, for WSA, CRO, and
SAV the observations in the highest VPD bins exhibit a downturn
towards more negative response, which cannot be captured by the
functional form of the sign term (Figure
\ref{idealized_sign}). Therefore, to explain the observations one of
the variables in the sign term must change at extreme VPD (upper
10-20\%). The most likely candidate is $uWUE$, which we have assumed
constant to meet our goal of developing intuition for leading order
behavior, but might be expected to decrease at extremely low SWC, and should be correlated with high VPD in very dry environments
through sensible heating increases \citep{Gentine_2016}. The theory
for EBF exhibits similar limitations as for ENF, WSA, and SAV, except
a larger portion of observations are biased negative as compared to
the theory ($\approx$ 35\%, Figure \ref{test_sign}). However, in
general, and specifically for non-extreme VPD (VPD < $\approx$70-90th
percentiles), the theory matches the observations for the tree and
savannah plant types well.

@@

While the above discussion shows that our theory has some limitations
when applied to some PFTs, especially for grasslands, it does well for
DBF, MF and CSH PFTs, and captures the response at non-extreme VPD
(VPD < $\approx$ 70th-90th percentiles) for
ENF, EBF, CRO, WSA, and SAV. In general, the leading order behavior
observed in the data is captured by the theory. Departures between our
theory and observations, specifically at high VPD, could be explained
and conceptualized with shifts in $g_1$ and/or $uWUE$ due to
site-specific plant type variability (e.g. more arid-adapted ecosystems
at more arid sites), or temporal variability for some environmental
conditions (e.g. decreases in $uWUE$ at extremely low soil
moisture). To focus on general behavior and develop intuition for
PFT-scale response, we have ignored these sources of variability in
the present analysis. However, any site-specific or temporal plant
functional variability that can be conceptualized with shifts to $g_1$
and $uWUE$ can be analyzed with our framework. This opens a door to future analyses in which plant behavior in anomalous conditions
can be explained and analyzed using Equation \ref{et}.

@@

Therefore, with the caveat that some of the signal presented in Figure
\ref{data_scatter} may in fact be noise, we can interpret the observed
distribution of ET response to VPD. In general, the observed response
matches the intuitive theory. ET response to VPD shifts towards
positive values as VPD increases (atmospheric demand dominating). CRO
exhibit the highest occurrence of positive ET response, and the
observations confirm that CRO are the most water intensive. CSH are
the most water conservative, with a strong negative response. DBF,
EBF, ENF, MF, SAV and WSA are also water conservative, but show some
occurrence of positive ET response to VPD, particularly at higher
observed VPD, as the theory predicts. GRA, while generally water
conservative, does not match the theory well, with increasing frequency of
\textit{negative} response at high VPD. This is as expected, given
previous discussion on how we might expect more inter-site and inter-year variability for GRA (Section \ref{testing}).

@@

Indeed, for all PFTs, there is some slight dependence of
$\sigma \cdot uWUE$ on SWC, especially at low SWC (Figure
\ref{swc_boxplot}). The portion of observations for which our theory
is biased by SWC-limitations varies by PFT, due to the nonlinear
threshold at which soil moisture availability limits plant
function. For CRO, MF and DBF, soil water content only matters for
about the lowest 5\% of observations (each box is 5\% of observations
in Figure \ref{swc_boxplot}). And, CRO, relative to DBF, has a very
weak dependence on soil moisture, which is reflective of the high
likelihood that CRO sites are optimally irrigated and not water
stressed, suggesting that observed departures from the theory for CRO
are due to other factors (i.e. different photosynthesis pathways; C3/C4) than soil
moisture. For ENF, EBF, SAV, and WSA, systematic SWC-induced biases in
$\sigma \cdot uWUE$ emerge in about the lowest 20-30\% of SWC
conditions, although large variability in the SWC-$\sigma \cdot uWUE$
relationship hampers interpretation for WSA. CSH presents a special
case: the limited number of sites (2) preclude comment on the
exact PFT-wide relationship.

@@

The observed dependence of $\sigma \cdot uWUE$ on SWC for GRA would
explain the deficiencies of our theory compared to the observations in
Figure \ref{idealized_sign}, specifically the trend back towards
negative soil moisture at high VPD. Aforementioned feedbacks between
the land surface and the atmosphere, which are not accounted for due
to our focus on the one-way response of the land surface to
atmospheric conditions, would cause high VPD to be correlated with low
SWC \citep[][]{Gentine_2016, Berg_2016}. So, at high VPD observations
of low SWC are more likely, and this low SWC causes a lower uWUE
(Figure \ref{swc_boxplot}). The lower uWUE at low SWC/high VPD then
leads to the observed downturn towards decreasing ET with increasing
VPD, and the deviation between our theory (based on a constant uWUE
assumption) and observations. It is also perhaps not coincidental that
the portion of $\sigma \cdot uWUE$ affected by low soil moisture
observations is similar to the portion of observations that do not
match our theory at high VPD. Indeed, this may suggest that for all PFTs
except for GRA, coupling and feedbacks between SWC, VPD and plant
function are relatively rare.

@@

The theory described in Section \ref{sign_term} indicates that for a
given $uWUE$ and $g_1$, the ET dependence on VPD should be concave
upward, which is confirmed by eddy covariance data cross most PFTs. In other words, there should be some local minimum in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms  (e.g. aerodynamic
conductance, $\Delta$, $g_1$ and $uWUE$) are held fixed. This result warrants further
investigation, because to our knowledge no one has derived the
theoretical ecosystem-scale relationship between ET and VPD while
controlling for other environmental conditions. In particular, from
personal communication, there is an apparent lack of consensus over
whether the shape of the ET-VPD curve should be concave upward (our
result) or concave downward in the absence of dramatic water stress. Given that understanding the ET-VPD
relationship of the one-way plant response is fundamental to
hypothesizing about any feedbacks between the land surface and the
atmosphere, we analyze why our derived ET-VPD relationship is concave
upward, particularly with respect to the exponent of VPD dependence in
$uWUE$ and the Medlyn unified stomatal conductance model.

There is a theoretical basis for the square root VPD dependence in
both the stomatal conductance model and $uWUE$ based on the assumption
that stomata behave to maximize carbon gain while minimizing water
loss, which observations also generally support \citep{Lloyd_1991,
  MEDLYN_2011, Lin_2015, Zhou_2014, Zhou_2015, Medlyn_2017}. However,
some purely empirical results that fit the exponent of the VPD
dependence to data have shown that it may vary slightly from 1/2,
suggesting that stomata, as well as ecosystem-scale quantities based
on stomata theory, may not always function optimally \citep{Zhou_2015,
  Lin_2018}. Specifically with regards to $uWUE$, one would not expect
that this ecosystem scale WUE quantity will respond to VPD exactly
analogously to stomata. Direct soil evaporation's contributions to ET
should shift the exponent of the VPD dependence, especially for
conditions of low GPP when we would expect a systematically larger
portion of direct soil evaporation contributions to ET, because we
would also expect lower amounts of transpiration. \citet{Zhou_2015}'s
results corroborate this: they found a mean empirically fit
exponential VPD dependence of 0.55, varying slightly from the
theoretically optimal value of of 0.5 for AmeriFlux
sites. Interpreting \citet{Lin_2018}'s results, which also show
variance in the empirical exponent of the VPD dependence of the
stomatal conductance model, is more difficult as \citet{Lin_2018} do
not handle GPP/A dependence of stomatal conductance in a directly
analogous manner to the optimal theory in \citet{MEDLYN_2011} and
\citet{Medlyn_2017}. Regardless, given that these recent results on
the relationship between VPD, GPP, and ET \citep{MEDLYN_2011,
  Zhou_2014, Zhou_2015, Medlyn_2017} form the backbone of our analysis
and are what allowed us to derive an explicit ET expression for the
first time (Equation \ref{et}), we will analyze if and how assumptions
about the exponent of the VPD dependence impacts the shape of the
ET-VPD dependence. This analysis is also important to understand whether the choice of stomatal conductance model alters the fundamental behavior of the ET-VPD relatioships, as many commonly used models  utilize a VPD exponent other than the 1/2 suggested by optimal theory \citep[e.g. ][ which uses an exponent of 1]{Leuning_1990}.

By introducing $n$ and $m$ we can free our stomatal
conductance model from assumptions about VPD dependence:
\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_*}{VPD^m}\right) \frac{*WUE \; ET}{c_a \; VPD^n},
    \label{m_n}
  \end{equation}
\end{linenomath*}
where:
\[*WUE = \frac{GPP}{ET}VPD^n,\] and $g_*$ is a generic slope parameter
of units $VPD^m$. To determine how the exponent $n$ and $m$ alter the
shape of the ET-VPD dependence we find the roots of the second
derivative of ET, using Equation \ref{m_n} for stomatal conductance
($g_s$, with respect to VPD: \input{d2_solutions.tex} With this
result we have defined the family of curves separating concave up from
concave down ET solutions (Figure \ref{concave}). These curves are
only functions of the exponent of the VPD dependence and a quantity we
call non-dimensional VPD ($VPD^m/g*$). Several important
relations reveal themselves from Equation \ref{curves}:

@@

\begin{enumerate}
  \item For optimal behavior (n, m = 1/2) the ET-VPD curve will be
    concave up regardless of the magnitude of the plant constants
    $g_1$ and $uWUE$. Therefore, the general concave up nature of our
    results, given an assumption of optimal behavior, is insensitive
    to plant type.
  \item For all physically possible exponents of VPD dependence ($n,
    m$), whether the solution is concave up or concave down does not
    depend on $uWUE$.
  \item In general, increasing the exponent of VPD dependence
    increases the likelihood of a concave up result. Additionally, as
    the exponent of VPD dependence increases from the optimum value of
    1/2, whether the curve is concave upward or concave downward
    becomes a function of the plant specific slope parameter $g_*$,
    through non-dimensional VPD ($VPD^m/g_*$). Because the exponent of
    the VPD dependencies is capable of altering the fundamental shape
    of ET-VPD dependence, future research investment in understanding
    the exact VPD dependence of stomatal conductance, and further
    reconciliation of empirical and theoretical stomatal and
    ecosystem behavior should be prioritized.
\end{enumerate}
While it is possible that in the future some other form of
VPD dependence is derived, at present \cite{MEDLYN_2011} and
\cite{Zhou_2014} firmly established n=m=1/2 as the most likely
candidate given current theory and empirical data. Additionally, we
argue that a concave up result matches physical intuition more than a
concave down result. Plants must maintain nutrient and sugar transport
through the phloem and xylem. To accomplish this, stomata must remain
slightly open \citep{De_2013, Nikinmaa_2013, Ryan_2014}. Furthermore,
even if complete stomatal closure were possible, cuticular water loss
and [at the ecosystem-scale] direct soil evaporation are still sources
of ET which increase with VPD, independent of stomatal
closure. Therefore, in the limit as VPD becomes large and we assume
plants are exercising all strategies to reduce ET, any further increase
in VPD should result in an increase in ET through cuticular water loss
and/or direct soil evaporation. This inevitable transition from
conditions when stomata respond strongly to VPD to conditions when
stomata response is asymptoting towards full closure would cause a
concave up ET-VPD curve, which is matched by the theory. In short,
plant response becomes more limited as VPD increases, while
atmospheric demand monotonically increases with VPD, leading to the
 result that atmospheric demand dominates plant response when
atmospheric demand is high.

@@

Lastly, our paper builds important intuition for how plants will
respond to VPD perturbations. We show that given optimal stomatal
function and fixed environmental conditions, the ET-VPD dependence is
theoretically concave up, with ET increasing with increasing VPD as
VPD increases past some critical value (Table \ref{vpd_crit}). However
future research should focus on fully understanding the form of
stomatal VPD dependence, as this result is sensitive to the exponent of
VPD dependence, which we currently believe is 1/2 \citep{MEDLYN_2011,
  Zhou_2014}. Our results are applicable to understanding the impact
of  expected increases in VPD
induced by global change.  Plant physiological responses to direct
CO$_2$ effects \citep[e.g.,][]{Swann_2016, Lemordant_2018} receives more attention
than physiological response to indirect effects like increased
VPD. Here, we provide broad PFT-focused results showing a likely
decrease in ET in response to positive VPD perturbations (atmospheric
drying), which is consistent with recent observational analysis
\citep[e.g.,][]{Rigden_2017}. Feedbacks between the land and the
atmosphere may alter the net response to a long-timescale global VPD perturbation,
but our focus on the one way plant response to a VPD perturbation in
the atmospheric boundary layer is an important first step to
disentangling such feedbacks, both in observations and model
simulations of the present and future. By removing Penman Monteith's
dependence on implicit relationships between GPP, VPD, and ET, we
allow for explicit, accurate future analysis of plant-VPD feedbacks
in the atmospheric boundary layer (Equation \ref{et}). Our approach
can be extended to examine varying plant response to more nuanced
consideration of plant type and climate. Any plant physiological
heterogeneity or feedback that can be conceptualized with shifts in $g_1$
\citep[e.g.][]{Lin_2015, Medlyn_2017} and/or uWUE
\citep[e.g.][]{Zhou_2014} are representable within our framework, which
opens the door for a hierarchy of more sophisticated climate- and
plant-specific analysis of ET sensitivity to environmental variables
(including VPD). We argue that such simplified conceptual frameworks
are critical tools for disentangling land-atmosphere feedbacks at
various scales, from diurnal to seasonal and beyond, and to
characterize ET response in a warmer, atmospherically drier, and
enriched CO$_2$ world.
